---
title: "GPX Cleaning"
author: "wolf zinke"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = getwd())
```

# define parameters

```{r}
filename  = "Afternoon_Hike.gpx"
act_typ   = 'w'  # cycling (c), running (r), walking/hiking (w)
lgnd_algn = 'bl' # top-left (tl), top-right (tr), bottom-left (bl), bottom-right (br)
mpsrc     = 'stamen'
mptyp     = 'watercolor'

# cleaning parameters

### walking
if(act_typ == 'w'){
  spdthr   = 8  # km/h - threshold to identify track points that exceed the expected distance
  sm_tm    = 1 # [minutes] - time window for loess smoothing
  ResThr   = 0.00001  # residual threshold after loess filtering
  mapzoom  = 14
}

### running
if(act_typ == 'r'){
  spdthr   = 15  # km/h - threshold to identify track points that exceed the expected distance
  sm_tm    = 0.5 # [minutes] - time window for loess smoothing
  ResThr   = 0.00005  # residual threshold after loess filtering
  mapzoom  = 14
}

### cycling
if(act_typ == 'c'){
  spdthr   = 60  # km/h - threshold to identify track points that exceed the expected distance
  sm_tm    = 0.5 # [minutes] - time window for loess smoothing
  ResThr   = 0.00001  # residual threshold after loess filtering
  mapzoom  = 13
}

# legend alignment
if(lgnd_algn == 'bl'){
  lgndpos  = c(.025, .025)
  lgndjust = c("left", "bottom")
}

if(lgnd_algn == 'tl'){
  lgndpos  = c(.025, .975)
  lgndjust = c("left", "top")
}

if(lgnd_algn == 'br'){
  lgndpos  = c(.975, .025)
  lgndjust = c("right", "bottom")
}

if(lgnd_algn == 'tr'){
  lgndpos  = c(.975, .975)
  lgndjust = c("right", "top")
}
```

Try to remove outliers from a gpx track and smooth the data.

some inspirations came from these resources:
  * https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/vignettes/gpx-files.Rmd
  * https://bookdown.org/robinlovelace/geocompr
  * https://padpadpadpad.github.io/post/animate-your-strava-activities-using-rstrava-and-gganimate
  * https://rspatial.org/raster/spatial/index.html
  * https://www.r-bloggers.com/stay-on-track-plotting-gps-tracks-with-r
  * https://digital-geography.com/gpx-overview-r-function-create-overview-gpx-files-using-leaflet-rgooglemaps

# load required packages
  
Load all packages required for the analysis. If these packages are not available, install them on the fly with a `install.packages(<packagesname>)` call.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
#Pckglst  = c("ggplot2", "plotKML", "maptools", "XML", "rgdal", "chron", "spatstat",
#             "dplyr", "purrr", "pgirmess")

Pckglst  = c("ggplot2", "maptools", "XML", "rgdal", "spatstat", "png",
             "dplyr", "purrr", "sp", "ggmap", "raster", "viridis", "stringi")

for(cPckg in Pckglst){
  if(!require(cPckg , character.only=TRUE)){
      install.packages(cPckg)
      library(cPckg , character.only = TRUE)
  }  
}
``` 

# read in gpx file

## define interpolation function

```{r}
Track_ip = function(trk,remark, smpldur=NULL){

if(is.null(smpldur)){ smpldur = round(median(diff(trk$Duration)), 10) }
  
nDuration = seq(from=0, to=max(trk$Duration), by=smpldur)

Lon_ip = approxfun(trk$Duration, trk$Longitude, rule=2)
Lat_ip = approxfun(trk$Duration, trk$Latitude,  rule=2)
Ele_ip = approxfun(trk$Duration, trk$Elevation, rule=2)

nLon = Lon_ip(nDuration)
nLat = Lat_ip(nDuration)
nEle = Ele_ip(nDuration)

Dist = spDists(x=cbind(nLon, nLat), longlat=TRUE, segments=TRUE)
Dist = c(0, sqrt(Dist^2 + (diff(nEle)/1000)^2) )

Spd  = c(Dist / smpldur)   # km/h

Range = cumsum(Dist)

lTm = as.difftime(nDuration, units="hours") + trk$Time[1]

trk_df = data.frame(Latitude=nLat, Longitude=nLon, Elevation=nEle, Time=lTm, 
                   Duration=nDuration, Distance=Dist, Speed=Spd, Range=Range)
return(trk_df)
}

```

```{r}
# from https://stackoverflow.com/questions/36377252/import-gpx-track-using-the-xml-library

gpx_dt = filename %>%
  xmlTreeParse(useInternalNodes = TRUE) %>%
  xmlRoot %>%
  xmlToList %>%
  (function(x) x$trk) %>%
  (function(x) unlist(x[names(x) == "trkseg"], recursive = FALSE)) %>%
  map_df(function(x) as.data.frame(t(unlist(x)), stringsAsFactors=FALSE))

# names(gpx_dt) # verify attributes labels
names(gpx_dt) = c("Elevation", "Time", "Latitude", "Longitude")

gpx_dt$Longitude = as.numeric(gpx_dt$Longitude)
gpx_dt$Latitude  = as.numeric(gpx_dt$Latitude)
gpx_dt$Elevation = as.numeric(gpx_dt$Elevation)
gpx_dt$Time      = as.POSIXct(gpx_dt$Time, tz="GMT", format="%Y-%m-%dT%H:%M:%OS")
gpx_dt$Duration  = as.numeric(difftime(gpx_dt$Time, gpx_dt$Time[1]), units="hours")

Npts = length(gpx_dt$Longitude)
sTm = gpx_dt$Time[1]
smplitv = round(median(diff(gpx_dt$Duration)),8)
```


# identify outliers and create new gpx data



## speed criterion

```{r}
gpx_cl = Track_ip(gpx_dt)

Nol = 1
while(Nol > 0){
  
  ol_p = which(gpx_cl$Speed > spdthr)
  Nol  = length(ol_p)
    
  gpx_cl[ol_p, ] = NA

  gpx_cl = droplevels(subset(gpx_cl, is.finite(gpx_cl$Longitude)))

  gpx_cl = Track_ip(gpx_cl)
}

plot(gpx_dt$Longitude, gpx_dt$Latitude, type='l')
lines(gpx_cl$Longitude, gpx_cl$Latitude, col='red')

plot(gpx_dt$Duration, gpx_dt$Elevation, type='l')
lines(gpx_cl$Duration, gpx_cl$Elevation, col='red')

gpx_ip = Track_ip(gpx_cl, smpldur=smplitv/2)
```


## loess residuals

```{r}
# running median residuals
# rMed_win = 151 # window size of data points used for the running median
#
# rLat = gpx_ip$Latitude  - runmed(gpx_ip$Latitude,  rMed_win)
# rLon = gpx_ip$Longitude - runmed(gpx_ip$Longitude, rMed_win)

# loess residuals
WinDur =   sm_tm / 60 # loess Window duration in hours (walking)

NWinPts = WinDur/median(diff(gpx_ip$Duration))
Lspan   = NWinPts / length(gpx_ip$Latitude)

rLon = loess(gpx_ip$Latitude  ~ gpx_ip$Duration, span=Lspan)$residuals
rLat = loess(gpx_ip$Longitude ~ gpx_ip$Duration, span=Lspan)$residuals
rEle = (loess(gpx_ip$Elevation ~ gpx_ip$Duration, span=Lspan)$residuals)/1000

rDist = sqrt(rLat^2 + rLon^2 + rEle^2)
# remove outliers

# ol = which(rDist > 10*mad(rDist))
ol = which(rDist > ResThr)

gpx_ol_rm = gpx_ip
gpx_ol_rm[ol, ] = NA

gpx_ol_rm = droplevels(subset(gpx_ol_rm, is.finite(gpx_ol_rm$Longitude)))
gpx_ol_rm = Track_ip(gpx_ol_rm)

# visualize
plot(gpx_ip$Longitude, gpx_ip$Latitude, type='l')
lines(gpx_ol_rm$Longitude, gpx_ol_rm$Latitude, col='red')

plot(gpx_ip$Duration, gpx_ip$Elevation, type='l')
lines(gpx_ol_rm$Duration, gpx_ol_rm$Elevation, col='red')
```

# smooth the data
```{r}
WinDur = sm_tm / 120 # loess Window duration in hours

gpx_sm = gpx_ol_rm

NWinPts = WinDur/median(diff(gpx_ol_rm$Duration))
Lspan   = NWinPts / length(gpx_ol_rm$Latitude)

gpx_sm$Longitude = loess(Longitude ~ Duration, data=gpx_ol_rm ,span=Lspan)$fitted
gpx_sm$Latitude  = loess(Latitude  ~ Duration, data=gpx_ol_rm, span=Lspan)$fitted
gpx_sm$Elevation = loess(Elevation ~ Duration, data=gpx_ol_rm, span=Lspan)$fitted

gpx_sm = Track_ip(gpx_sm)


plot(gpx_dt$Longitude, gpx_dt$Latitude, type='l')
lines(gpx_ol_rm$Longitude, gpx_ol_rm$Latitude, col='red')
lines(gpx_sm$Longitude, gpx_sm$Latitude, col='green')

plot(gpx_dt$Duration,gpx_dt$Elevation, type='l')
lines(gpx_dt$Duration,gpx_dt$Elevation, col='red')
lines(gpx_sm$Duration,gpx_sm$Elevation, col='green')
```

# write processed GPX file

## define Write GPX
from: 
https://stackoverflow.com/questions/54726758/merging-multiple-gpx-files-into-a-single-gpx-file-with-multiple-tracks

```{r}
writeGPX = function(gpx_trk, file="file.gpx"){
  o = c('<gpx version="1.1" creator="R">','<trk>','<trkseg>')
  o = c(o, paste('<trkpt lat="', gpx_trk$Latitude,'" lon="', gpx_trk$Longitude,'"><time>',
           paste("<ele>", gpx_trk$Elevation,"</ele>",sep=""),
           paste(gsub(' ','T', as.character(gpx_trk$Time)), 'Z', sep=''),
           '</time></trkpt>', sep=''))
  
  o = c(o, '</trkseg>', '</trk>', '</gpx>')
  cat(o, file=file, sep='\n')
}
```

## Write data
```{r}

writeGPX(gpx_sm, file="gpx_cleaned.gpx")
```

# plot speed profile

```{r}
# get the map
bbox = make_bbox(gpx_sm$Longitude, gpx_sm$Latitude)
b1   = get_map(bbox, maptype=mptyp, source=mpsrc, force=TRUE, zoom=mapzoom)


gpx_pl = gpx_sm

mx_spd = quantile(gpx_pl$Speed, 0.99)
gpx_pl$Speed[gpx_pl$Speed>mx_spd] = mx_spd

ggmap(b1) + 
  geom_point(data=gpx_pl, 
             aes(Longitude, Latitude, col=Speed), size=1.5, alpha=0.5) +
  scale_colour_viridis(option='inferno', limits = c(0, mx_spd)) + theme_void() + 
  theme(legend.position=lgndpos, legend.justification=lgndjust,
        legend.title = element_text(size=10),
        legend.text  = element_text(size=8))

# get plot area instead of bbox

hgth = abs(diff(bbox[c(2,4)]))
wdth = abs(diff(bbox[c(1,3)]))

wdthscl = 10
hgthscl = wdthscl * (hgth/wdth)

ggsave('speedmap.png', width=wdthscl, height=hgthscl)
```
