---
title: "GPX Cleaning"
author: "wolf zinke"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = getwd())
```

Try to remove outliers from a gpx track and smooth the data.

some inspirations came from these resources:
  * https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/vignettes/gpx-files.Rmd
  * https://bookdown.org/robinlovelace/geocompr
  * https://rspatial.org/raster/spatial/index.html
  * https://www.r-bloggers.com/stay-on-track-plotting-gps-tracks-with-r
  * https://digital-geography.com/gpx-overview-r-function-create-overview-gpx-files-using-leaflet-rgooglemaps

# load required packages
  
Load all packages required for the analysis. If these packages are not available, install them on the fly with a `install.packages(<packagesname>)` call.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
#Pckglst  = c("ggplot2", "plotKML", "maptools", "XML", "rgdal", "chron", "spatstat",
#             "dplyr", "purrr", "pgirmess")

Pckglst  = c("ggplot2", "maptools", "XML", "rgdal", "spatstat",
             "dplyr", "purrr", "sp")

for(cPckg in Pckglst){
  if(!require(cPckg , character.only=TRUE)){
      install.packages(cPckg)
      library(cPckg , character.only = TRUE)
  }  
}
``` 


# read in gpx file

```{r}
# from https://stackoverflow.com/questions/36377252/import-gpx-track-using-the-xml-library
filename = "EM_Lunch_Walk.gpx"

gpx_dt <- filename %>%
  xmlTreeParse(useInternalNodes = TRUE) %>%
  xmlRoot %>%
  xmlToList %>%
  (function(x) x$trk) %>%
  (function(x) unlist(x[names(x) == "trkseg"], recursive = FALSE)) %>%
  map_df(function(x) as.data.frame(t(unlist(x)), stringsAsFactors=FALSE))

# names(gpx_dt) # verify attributes labels
names(gpx_dt) = c("Elevation", "Time", "Latitude", "Longitude")

gpx_dt$Longitude = as.numeric(gpx_dt$Longitude)
gpx_dt$Latitude  = as.numeric(gpx_dt$Latitude)
gpx_dt$Elevation = as.numeric(gpx_dt$Elevation)
gpx_dt$Time      = as.POSIXct(gpx_dt$Time, tz="GMT", format="%Y-%m-%dT%H:%M:%OS")
gpx_dt$Duration  = as.numeric(difftime(gpx_dt$Time, gpx_dt$Time[1]), units="hours")

Npts = length(gpx_dt$Longitude)
sTm = gpx_dt$Time[1]
```

# Determine Distance and Speed
```{r}
Dist     = spDists(x=cbind(gpx_dt$Longitude, gpx_dt$Latitude), longlat=TRUE, segments = TRUE) 
interval = diff(gpx_dt$Duration)
smpldur  = round(median(interval)/2, 10)

Speed = c(0, Dist / interval)   # km/h
```

# identify outliers and create new gpx data

```{r}
# running median residuals
# rMed_win = 151 # window size of data points used for the running median
#
# rLat = gpx_dt$Latitude  - runmed(gpx_dt$Latitude,  rMed_win)
# rLon = gpx_dt$Longitude - runmed(gpx_dt$Longitude, rMed_win)

# loess residuals
ResThr = 0.0001
WinDur = 1 / 60 # loess Window duration in hours

NWinPts = WinDur/smpldur
Lspan   = NWinPts / Npts

rLon = loess(gpx_dt$Latitude  ~ gpx_dt$Duration, span=Lspan)$residuals
rLat = loess(gpx_dt$Longitude ~ gpx_dt$Duration, span=Lspan)$residuals

rDist = sqrt(rLat^2 + rLon^2)

# remove outliers

# ol = which(rDist > 10*mad(rDist))
ol = which(rDist > ResThr)

gpx_ol_rm = gpx_dt
gpx_ol_rm[ol, ] = NA

gpx_ol_rm = droplevels(subset(gpx_ol_rm, is.finite(gpx_ol_rm$Longitude)))

# define interpolation functions
nDuration = seq(from=0, to=max(gpx_dt$Duration), by=smpldur)

Lon_ip = approxfun(gpx_ol_rm$Duration, gpx_ol_rm$Longitude, rule=2)
Lat_ip = approxfun(gpx_ol_rm$Duration, gpx_ol_rm$Latitude,  rule=2)
Ele_ip = approxfun(gpx_ol_rm$Duration, gpx_ol_rm$Elevation, rule=2)

nLon = Lon_ip(nDuration)
nLat = Lat_ip(nDuration)
nEle = Ele_ip(nDuration)

# visualize
plot(gpx_dt$Longitude, gpx_dt$Latitude)
lines(nLon, nLat, col='red')

plot(gpx_dt$Duration,gpx_dt$Elevation)
lines(nDuration,nEle, col='red')
```

# smooth the data
```{r}
WinDur = 2 / 60 # loess Window duration in hours

NWinPts = WinDur/smpldur
Lspan = NWinPts / length(nDuration)

lLon = loess(nLon ~ nDuration, span=Lspan)$fitted
lLat = loess(nLat ~ nDuration, span=Lspan)$fitted
lEle = loess(nEle ~ nDuration, span=Lspan)$fitted

plot(gpx_dt$Longitude, gpx_dt$Latitude)
lines(nLon, nLat, col='red')
lines(lLon, lLat, col='green')

plot(gpx_dt$Duration,gpx_dt$Elevation)
lines(nDuration,nEle, col='red')
lines(nDuration,lEle, col='green')
```

# write processed GPX file

## define Write GPX
from: 
https://stackoverflow.com/questions/54726758/merging-multiple-gpx-files-into-a-single-gpx-file-with-multiple-tracks

```{r}
writeGPX = function(Time, Longitude, Latitude, Elevation, file="file.gpx"){
  o = c('<gpx version="1.1" creator="R">','<trk>','<trkseg>')
  o = c(o, paste('<trkpt lat="', Latitude,'" lon="', Longitude,'"><time>',
           paste("<ele>", Elevation,"</ele>",sep=""),
           paste(gsub(' ','T', as.character(Time)), 'Z', sep=''),
           '</time></trkpt>', sep=''))
  
  o = c(o, '</trkseg>', '</trk>', '</gpx>')
  cat(o, file=file, sep='\n')
}
```

## Write data
```{r}
lTm = as.difftime(nDuration, units="hours") + sTm

writeGPX(lTm, lLon, lLat, lEle, file="gpx_cleaned.gpx")
```


# just some plots to play with


```{r}
#gpx_dt = readGPX(filename)
#gpx_raw <- readOGR(dsn = filename)
#pfile <- htmlTreeParse(filename, useInternalNodes = T)

gpx.raw = readGPS(i = "gpx", f = filename, type="w")
(layers = ogrListLayers(filename))
gp1 = readOGR(filename, layer=layers[5])
gt1 = readOGR(filename, layer=layers[3])
```


```{r}
plot(gp1)
plot(gt1, add = T, col = "red")
```

```{r}
plot(gp1$track_seg_point_id, gp1$ele )
```

```{r}
gp1.ppp = as(SpatialPoints(gp1), "ppp")
plot(density(gp1.ppp, adjust=0.5))
plot(gt1, add = T, col = "green")
```



